The effects of multi environmental exposures on Tuberculosis Incidence in South East Asian and African countris

### 2025-02-23 code/analysis-code/statistical-analysis.qdm.qmd
Add some important things

library(here)
library(knitr)
install.packages("kableExtra")
library(kableExtra)
install.packages("GGally")
library(GGally)
# Install and load the patchwork package
install.packages("patchwork")
library(patchwork)

@@

Load the dataset
```{r}

asia_africa_location <- here("data", "processed-data", "asia_africa.rds")
readRDS(asia_africa_location)

@@

@@

@@

# Step 2: Reshape the dataset for visualization
asia_africa_long <- asia_africa %>%
  select(-country, -region, -year) %>%  # Exclude categorical variables
  pivot_longer(cols = -tbi, names_to = "predictor", values_to = "value")

@@

```
Q-Q plot of residuals from the Poisson GLM suggests that the model is not fitting the data well. The deviation from the straight line, especially at higher quantiles (outliers labeled 78, 79, 80), indicates overdispersion—where the variance of tbi is larger than the mean, violating the Poisson assumption.

@@

```
since the result is very high, which should be close to 1. We need a model that can hanlde overdispersion

@@

- hdi (-7.642, p < 2e-16) Higher HDI is associated with lower tbi cases.
- pro_basicsanitation (0.00577, p = 0.0334)  Increased access to basic sanitation is linked to a slight increase in tbi.
- pro_basicwater (0.0318, p = 9.42e-16) → Higher access to basic water is positively associated with tbi.
- pro_safewater (-0.00856, p = 0.0414) → Increased safe water access reduces tbi.
- pro_cleanfuel (-0.005999, p = 0.0137) → More use of clean fuels is linked to lower tbi.
- mean_exposure_pm2.5 (-0.02698, p < 2e-16) → Higher PM2.5 exposure reduces tbi (unexpected; may need further investigation).

>>>>

library(splines)
library(here)
library(knitr)
install.packages("kableExtra")
library(kableExtra)
install.packages("GGally")
library(GGally)
# Install and load the patchwork package
install.packages("patchwork")
library(patchwork)

@@

Load the dataset

asia_africa_location <- here("data", "processed-data", "asia_africa.rds")
asia_africa <- readRDS(asia_africa_location)

@@

# Define independent variables (exclude 'tbi' and categorical variables)
independent_vars <- colnames(asia_africa)[!colnames(asia_africa) %in% c("tbi", "country", "region", "year")]

# Create scatter plots for each independent variable against tbi
plots <- lapply(independent_vars, function(var) {
  ggplot(asia_africa, aes_string(x = var, y = "tbi")) +
    geom_point() + 
    ggtitle(paste("TBI vs", var)) + 
    xlab(var) + 
    ylab("TBI") + 
    theme_minimal()
})

@@

Interpretation: The predictors have no linear relationship with Tuberculosis Incidence. Therefore, we would like to perform nonlinear relationship to decide further action and analysis.

***Check Nonlinear relationship***

# Reshape the dataset to long format
asia_africa_long <- asia_africa %>%
  pivot_longer(cols = c(pro_cleanfuel, pro_basicwater, pro_safewater, 
                        mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, 
                        pro_safesanitation, gdp, hdi),
               names_to = "Predictor", values_to = "Value")

# Plot nonlinear relationships using LOESS smoothing
nonlinear <- ggplot(asia_africa_long, aes(x = Value, y = tbi)) +
  geom_point(alpha = 0.5) +  # Scatter points
  geom_smooth(method = "loess", color = "red") +  # LOESS curve
  facet_wrap(~ Predictor, scales = "free_x") +  # Separate plots for each predictor
  theme_minimal() +
  labs(title = "Nonlinear Relationship between TBI and Predictors",
       x = "Predictor Value",
       y = "TBI")

nonrelationshiplot_location <- here( "code","analysis-code", "figures", "nonlinear.png") # to set up location for the pictures created 
ggsave(filename = nonrelationshiplot_location, plot=nonlinear, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(nonlinear)
```

Data transformation

library(dplyr)

asia_africa <- asia_africa %>%
  mutate(
    log_tbi = log(tbi + 1),
    log_gdp = log(gdp + 1),
    log_total_greenhouses = log(total_greenhouses + 1),
    log_pm2.5 = log(mean_exposure_pm2.5 + 1),
    sqrt_basicwater = sqrt(pmax(pro_basicwater, 0)), 
    sqrt_safewater = sqrt(pmax(pro_safewater, 0)),
    poly_cleanfuel = as.numeric(poly(pro_cleanfuel, 2, raw = TRUE)[, 1]),
    poly_basicsanitation = as.numeric(poly(pro_basicsanitation, 2, raw = TRUE)[, 1]),
    poly_safesanitation = as.numeric(poly(pro_safesanitation, 2, raw = TRUE)[, 1])
  )

```{r}
asia_africa_transformed <- asia_africa %>%
  dplyr::select(log_tbi, log_gdp, log_total_greenhouses, log_pm2.5, 
                poly_cleanfuel, poly_basicsanitation, poly_safesanitation,
                sqrt_basicwater, sqrt_safewater, hdi) %>%
  pivot_longer(cols = -log_tbi, names_to = "Predictor", values_to = "Value")

colnames(asia_africa)

# Plot after transformation
transformed <- ggplot(asia_africa_transformed, aes(x = Value, y = log_tbi)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +  # Use linear model for comparison
  facet_wrap(~ Predictor, scales = "free_x") +
  theme_minimal() +
  labs(title = "Transformed Predictor Relationships with Log(TBI)",
       x = "Transformed Predictor Value", y = "Log(TBI)")

transformedplot_location <- here( "code","analysis-code", "figures", "transformed.png") # to set up location for the pictures created 
ggsave(filename = transformedplot_location, plot=transformed, width = 12, height = 12, units = "in", dpi = 300) # save the pictures created 
plot(transformed)
```

After transforming variables. They look more linear. Now GLM will be performed

***GLMs***

```{r}
glm_model <- glm(log_tbi ~ log_gdp + log_total_greenhouses + log_pm2.5 +
                      poly_cleanfuel + poly_basicsanitation + poly_safesanitation +
                      sqrt_basicwater + sqrt_safewater + hdi,
                 data = asia_africa, family = gaussian)
summary(glm_model)

\*\*\*General Addictive Model (GAM)

gam_model <- gam(log_tbi ~ s(log_gdp) + s(log_total_greenhouses) + s(log_pm2.5) +
                      s(poly_cleanfuel) + s(poly_basicsanitation) + s(poly_safesanitation) +
                      s(sqrt_basicwater) + s(sqrt_safewater) + s(hdi),
                 data = asia_africa, family = gaussian)
summary(gam_model)

Generalized Additive Model (GAM) explains 94.5% of the deviance in log-transformed tuberculosis incidence (Log(TBI)), which suggests an excellent fit. Below is a step-by-step breakdown of the interpretation.

Overall Model Performance Adjusted R² (R-sq.(adj) = 0.917) - This means 91.7% of the variability in Log(TBI) is explained by the predictors. - A very strong model since it explains most of the variance.

Deviance Explained = 94.5%

-   Similar to R² but for GAM, indicating excellent model fit.

Generalized Cross-Validation (GCV) = 0.0685

-   Lower GCV indicates better predictive performance.
-   This suggests that the model generalizes well to new data.

![](images/clipboard-4125844072.png)

Now, I want to remove insignificant variables

```{r}
gam_optimized <- gam(log_tbi ~ s(log_gdp) + s(log_total_greenhouses) + s(log_pm2.5) + 
                         s(poly_cleanfuel) + s(sqrt_basicwater) + 
                         s(sqrt_safewater) + s(hdi),
                     data = asia_africa, family = gaussian)
summary(gam_optimized)

Comparing GAM models

```{r}
AIC(gam_model, gam_optimized)
BIC(gam_model, gam_optimized)

Visualize significant predictors

```{r}
plot(gam_model, pages = 1, shade = TRUE)

```{r}
ggplot(data = asia_africa, aes(x = exp(log_tbi), y = exp(predict(gam_model, type = "response")))) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Actual vs. Predicted TBI",
       x = "Actual TBI", y = "Predicted TBI")

***Comparing the model***

```{r}
AIC(glm_model, gam_model)
BIC(glm_model, gam_model)
```

After comapring the models, based on AIC and BIC, GAM model is more appropriate

Check Paryial effects of Predictors

```{r}
plot(gam_model, pages = 1, shade = TRUE)

```{r}
predictions <- predict(gam_model, newdata = asia_africa, type = "response")
head(predictions)

Compare actual vs predicted values

```{r}
ggplot(data = asia_africa, aes(x = log_tbi, y = predictions)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Actual vs. Predicted Log(TBI)", x = "Actual Log(TBI)", y = "Predicted Log(TBI)")

asia_africa$predicted_tbi <- exp(predict(gam_model, newdata = asia_africa, type = "response"))

ggplot(asia_africa, aes(x = tbi, y = predicted_tbi)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Actual vs. Predicted TBI",
       x = "Actual TBI", y = "Predicted TBI")

The points are generally close to the line, indicating that the model is accurate.

@@

# Ensure it's a tibble
asia_africa <- as_tibble(asia_africa)

# Select variables (excluding categorical ones)
asia_africa_selected <- asia_africa %>%
  select(-all_of(c("country", "region", "year")))

# Reshape the dataset
asia_africa_long <- asia_africa_selected %>%
  pivot_longer(cols = -tbi, names_to = "predictor", values_to = "value")

# Print first few rows
print(head(asia_africa_long))

# Step 2: Reshape the dataset for visualization
asia_africa <- as_tibble(asia_africa)
asia_africa_long <- asia_africa %>%
  select(-c(country, region, year)) %>%  # Use `-c()` for multiple column exclusion
  pivot_longer(cols = -tbi, names_to = "predictor", values_to = "value")

@@

Q-Q plot of residuals from the Poisson GLM suggests that the model is not fitting the data well. The deviation from the straight line, especially at higher quantiles (outliers labeled 78, 79, 80), indicates overdispersion—where the variance of tbi is larger than the mean, violating the Poisson assumption.

@@

since the result is very high, which should be close to 1. We need a model that can hanlde overdispersion

@@

-   hdi (-7.642, p \< 2e-16) Higher HDI is associated with lower tbi cases.
-   pro_basicsanitation (0.00577, p = 0.0334) Increased access to basic sanitation is linked to a slight increase in tbi.
-   pro_basicwater (0.0318, p = 9.42e-16) → Higher access to basic water is positively associated with tbi.
-   pro_safewater (-0.00856, p = 0.0414) → Increased safe water access reduces tbi.
-   pro_cleanfuel (-0.005999, p = 0.0137) → More use of clean fuels is linked to lower tbi.
-   mean_exposure_pm2.5 (-0.02698, p \< 2e-16) → Higher PM2.5 exposure reduces tbi (unexpected; may need further investigation).

### 2025-02-23 products/manuscript/Manuscript.qmd
Add some important things

@@

*To get some further insight into your data, if reasonable you could
compute simple statistics (e.g. simple models with 1 predictor) to look
for associations between your outcome(s) and each individual predictor
variable. Though note that unless you pre-specified the outcome and main
exposure, any "p\<0.05 means statistical significance" interpretation is
not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

@@

*Use one or several suitable statistical/machine learning methods to
analyze your data and to produce meaningful figures, tables, etc. This
might again be code that is best placed in one or several separate R
scripts that need to be well documented. You want the code to produce
figures and data ready for display as tables, and save those. Then you
load them here.*

@@

>>>>

#| label: fig-6
#| fig-cap: "Nonlinear relationship"
#| echo: FALSE
knitr::include_graphics(here("code","analysis-code","figures","nonlinear.png"))

@@

After performing nonlinear model, data transformation used based on the shave or distribution of the plots from nonlinear relationship plots ( @fig-6 ) above. ( @fig-7 ) shows the data distribution after data transformation.

#| label: fig-7
#| fig-cap: "transformed data plots"
#| echo: FALSE
knitr::include_graphics(here("code","analysis-code","figures","transformed.png"))

@@

For the final model, Generalized linear models (GLMs) and Generalized Addictive Models (GAMs). I need to discuss with Dr. Handal to get advice about this.

I actually have performed both GLMs and GAMs, but I still need to learn both models in depth. Specially for GAM, I am very new in the models. Initially I performed GLMs since the Transformed plots show linears line, but aftrer comparing the GLMs and GAMs model, GAMs perfoming better.

@@

I will include reference for the sourse of data in the next step.
